Topological Equivariant Artist Model for Visualization Library Architecture

### 2022-07-28 paper/main.tex
equivariance section of rw

\subsection{Data Field Structure}
\label{sec:related-work:equivariance}
Along with preserving topology, visualization algorithms are expected to preserve the structure of the data fields, i.e. the column types in a table, when they translate data to visual representations. The mathematical term for this is equivariance and an equivariant map is a map that transforms input with structure to output with compatible structure. Bertin codified the idea that visual encoding of a data field should match the structure of the field; for example ordered data should be encoded as an ordered visual element or quantitative data as a quantitative visual element \cite{bertinSemiologyGraphicsDiagrams2011}. This structure mapping is necessary to ensure that the visual representations in graphs are not distorting the data\cite{tufteVisualDisplayQuantitative2001} and graphs are easier to understand when properties match \cite{norman_things_smart}. The expectation of matching structure is generalized as a mapping of a binary operator that acts on the data domain to a binary operator that acts on the graphic domain by Mackinlay\cite{mackinlayAutomaticDesignGraphical1987}. An invertible binary operator is the basis of Kindlemannn and Scheideggers's algebraic model of visualization\cite{kindlmannAlgebraicProcessVisualization2014}. They propose three specific types of equivariance: that the visualization should not change if the data representation (i.e. the data container) changes, that data maps unambiguously to visual elements \cite{ziemkiewiczEmbeddingInformationVisualization2009}, that changing the data should correspond to changes in the visualization in a perceptually significant manner. In \autoref{sec:atct:sheaves} we introduce a uniform abstract data representation layer, the expectation of unambiguous mappings is expressed in \autoref{sec:atct:xi}, and the definition of equivariance introduced in \autoref{sec:artist:equivariant:artist} includes the correspondence between data and visual changes. In this work, we formally specify the conditions necessary to build structure preserving components using category theory, because it provides a rich language for expressing the structure of objects, functions between objects, and the constraints that must hold for these functions to compose \cite{wielsManagementEvolvingSpecifications1998,yorgeyMonoidsThemeVariations}. A brief visualization oriented introduction to category theory is in Vickers et al \cite{vickersUnderstandingVisualizationFormal2013}, but they are applying category theory to semantic concerns about visualization design rather than library architecture.

@@

\subsection{Equivariance}
\label{sec:artist:equiv}
Equivariance in a visual context was formally defined as the preservation of a binary operator from one domain to another by Mackinlay\cite{mackinlayAutomaticDesignGraphical1987}. We generalize binary operations to actions $\dfunc \in \dfuncset$ on the sheaf $\sheaf_{\dbase, \dtotal}$.

>>>>

\subsection{Data Field Structure}
\label{sec:related-work:equivariance}
Along with preserving topology, visualization algorithms are expected to preserve the structure of the data fields, i.e. the columns in a table, when they translate data to visual representations. The field structure of data has traditionally been described using the measurment scales-i.e. nominal, ordinal, interval, ratio - which Steven's classified by their mathematical group structure\cite{stevensTheoryScalesMeasurement1946a} and others have formally expanded to include more types \cite{leaFormalizationMeasurementScale, thomasMathematizationNotMeasurement2014}. The expectation that visual transformation is one where the measurement structure be matched by the visual encoding was codified by Bertin \cite{bertinSemiologyGraphicsDiagrams2011}. An \textit{equivariant} map is the formal term for a function where the input and output have symmetric group structure. Given a group $G$ that acts on input $X$ and output $Y$ and a function $f: X \rightarrow Y$, the function $f$ is \textit{equivariant} if $f(g(x)) = g(f(x))$ for all $g$ in $G$ and for all $x$ in $X$\cite{nlab:equivariant}. For example, nominal measurements are permutable, meaning that they are acted on by elements of a permutation group. For the visual transformation to be equivariant, that nominal measurement must be mapped to a visual encoding that is also permutable, such as distinct marker shapes or color hues. Mackinlay\cite{mackinlayAutomaticDesignGraphical1987} more generally definines visualization transformations as \textit{homomorphic} maps. A map $f: X \rightarrow Y$ is a \textit{homomorphism} when $X$ and $Y$ have the same algebraic structure type and $f(x_1 \oslash x_2) = f(x_1) \oslash f(x_2)$ for every pair $x_1, x_2$ in $X$\cite{grimaldiDiscreteCombinatorialMathematics2006}. Mackinlay's construction encompasses non-invertable structures such as partial orderings. In \autoref{sec:artist:equivariant:data}, we will also describe the structure of the data in terms of the functions acting on the data.

Structure preservation is important because it means visualzations are easier to understand \cite{norman_things_smart} and ensures that the visual representations are not distortions of the data\cite{tufteVisualDisplayQuantitative2001}. Kindlemannn and Scheidegger\cite{kindlmannAlgebraicProcessVisualization2014} propose three specific types of structure preservation: that the visualization should not change if the data representation (i.e. the data container) changes, that data maps unambiguously to visual elements \cite{ziemkiewiczEmbeddingInformationVisualization2009}, that changing the data should correspond to changes in the visualization in a perceptually significant manner. In \autoref{sec:atct:sheaves} we introduce a uniform abstract data representation layer, the expectation of unambiguous mappings is expressed in \autoref{sec:atct:xi}, and the definition of equivariance introduced in \autoref{sec:artist:equivariant:artist} includes the correspondence between data and visual changes. In this work, we formally specify the conditions necessary to build structure preserving components using category theory, because it provides a rich language for expressing the structure of objects, functions between objects, and the constraints that must hold for these functions to compose \cite{wielsManagementEvolvingSpecifications1998,yorgeyMonoidsThemeVariations}. A brief visualization oriented introduction to category theory is in Vickers et al \cite{vickersUnderstandingVisualizationFormal2013}, but they are applying category theory to semantic concerns about visualization design rather than library architecture.

@@

\subsection{Equivariance}
\label{sec:artist:equiv}\note{our basis of structure preservation - is this even equivariance}
to actions $\dfunc \in \dfuncset$ on the sheaf $\sheaf_{\dbase, \dtotal}$.
